---
title: "YNP productivity and consumption"
author: "Jason Fridley"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=F)
```

### Update 6-21-23

Changes below are highlighted in yellow. These include:

1. Created a new ANPP variable (ANPP2) that ignores negatively monthly ANP values in the sum calculation. This did not change ANPP relationships with rainfall or elevation. I also added in mean temperature, calculated as the mean of monthly data.

2. Calculated ANPP that occurs after sample period 1 (ANPP.no1), and the proportion of ANPP that is produced in the first month (s1prop). Determined that if the first month of ANP is excluded, elevation is a significant predictor of ANP (negative relationship) and temperature is slightly negative. No effects of rainfall at the annual scale.

3. Added monthly predictor: %dead of total biomass (dead/total) each month, te treatment only. Added this predictor to the mANP model, and it did interact with rainfall.

4. Separated monthly models into all months vs. excluding month 1, including rainfall-temp-elevation interactions.

5. Changed percent consumed covariate (grazing intensity) to include cases of below 0 and above 100%, but this had no effect on the model outcome.

6. Changed the final model concerning long-term grazing effects on annual productivity, using the ANPP version that ignores negative monthly ANP values, and added site productivity (ANPP) as an explanatory variable, which had a large effect on Y.

---
This document summarizes monthly ANP (aboveground net productivity, g/m2) and monthly consumption by grazers (g/m2) from the YNP dataset, across sites and years. ANP is calculated as the difference in standing biomass between a temporary exclosure (te) plot and its initial biomass, estimated as the biomass of the grazed grassland for the prior sample period. Because observation intervals vary, productivity is calculated at the daily scale before adjusting to monthly values. Consumption (C) is calculated as the difference in biomass between grazed (gr) plots and exclosure (te) plots during the same period. There are as many as 12 replicate samples per site per period; calculations take the average value before taking differences. 

### YNP dataset

This is the version Doug sent on March 21, 2023. The May 31 version has some differences to data columns that need discussion. 

```{r}
setwd("C:\\Users\\fridley\\OneDrive - Clemson University\\academic\\projects\\YNP_Doug")
dat = read.csv("dat_03-21-23.csv")

#tidy up columns
dat$hits.pin = as.numeric(as.character(dat$hits.pin))
dat$mo.cons = as.numeric(as.character(dat$mo.cons))
dat$mo.stim = as.numeric(as.character(dat$mo.stim))
dat$soilc = as.numeric(as.character(dat$soilc))
dat$clay = as.numeric(as.character(dat$clay))
dat$silt = as.numeric(as.character(dat$silt))
dat$gs.temp = as.numeric(as.character(dat$gs.temp))

#create a %dead biomass column
dat$percDead = dat$dadbiom / (dat$live.biom + dat$dadbiom)
hist(dat$percDead)

```

The first step is to simplify the dataset, taking means (and standard errors) across replicates. I then created an output data frame for calculations, using all possible combinations of site, year, and sample period (1-7). The data are imbalanced so many NA values will be subsequently ignored.

```{r}

#aggregate data
library(doBy)
meanf = function(x) {mean(x,na.rm=T)}
sef = function(x) {sd(x,na.rm=T) / sqrt(length(x))}
datb = dat[is.element(dat$sample.,1:7),]
dat2 = summaryBy( totshtbiom + percDead ~ yr + site + sample. + trt + days.1 + juldate, data = datb, FUN=c(meanf,sef) )
str(dat2)

#create new matrix for ANP and C data
years = unique(dat2$yr)
site = unique(dat2$site)
sample = unique(as.factor(dat2$sample.))
newdat = expand.grid(site,years,sample); names(newdat) = c("site","year","sample")
newdat$dANP = NA #will be filled
newdat$dC = NA #will be filled
newdat$percDead = NA #will be filled
```

Calculations were performed separately for the first sample period (=1), where ANP and te biomass were assumed equivalent (converted to daily values). If the 'tr' sample was missing, I presumed this meant grazers were absent and thus ANP could be calculated with 'gr' treatment values. For all sample periods, whenever 'te' and 'gr' treatments were both available, I calculated consumption as the difference, corrected for differences in days of observation. Calculations were performed in two loops:

<div style="background-color: #FFFF00">Added: average percentage dead biomass in te treatments for a given sample period. </div>

```{r, warning=F, message=F}

#add dANP and dC for sample 1
for(i in 1:dim(newdat)[1]) {
  x = newdat[i,]
  yr = x$year; si = as.character(x$site); sa = x$sample
  if(sa!=1) next
  
  #calculate average percent dead biomass, just te treatments
  newdat$percDead[i] = mean(dat2$percDead.meanf[dat2$yr==yr&dat2$site==si&dat2$sample.==1&dat2$trt=="te"],na.rm=T)
  
  #calculate daily ANP for sample period 1, using just te biomass
  biomass.mean.x = dat2$totshtbiom.meanf[dat2$yr==yr&dat2$site==si&dat2$sample.==1&dat2$trt=="te"]
  days.x = dat2$days.1[dat2$yr==yr&dat2$site==si&dat2$sample.==1&dat2$trt=="te"]
  if(length(days.x)==0) days.x = 28 #if missing assume 28 d
  #if there is no te treatment, use gr instead (presumed no grazers around)
  if(length(biomass.mean.x)==0) biomass.mean.x = dat2$totshtbiom.meanf[dat2$yr==yr&dat2$site==si&dat2$sample.==1&dat2$trt=="gr"]
  if(length(biomass.mean.x)==0) next
  newdat$dANP[i] = biomass.mean.x/days.x
  
  #calculate consumption: compare current te value to current gr value
  biomass.mean.x = dat2$totshtbiom.meanf[dat2$yr==yr&dat2$site==si&dat2$sample.==1&dat2$trt=="te"]
  if(length(biomass.mean.x)==0) next
  grbiomass.mean.x = dat2$totshtbiom.meanf[dat2$yr==yr&dat2$site==si&dat2$sample.==1&dat2$trt=="gr"]
  if(length(grbiomass.mean.x)==0) next
  newdat$dC[i] = (biomass.mean.x - grbiomass.mean.x) / days.x
}

#add dANP and dC for samples 2-7
for(s in 2:7) { #loop over samples
  for(i in 1:dim(newdat)[1]) {
    x = newdat[i,]
    yr = x$year; si = as.character(x$site); sa = x$sample
    if(sa!=s) next
    
    #calculate average percent dead biomass, just te treatments
    newdat$percDead[i] = mean(dat2$percDead.meanf[dat2$yr==yr&dat2$site==si&dat2$sample.==s&dat2$trt=="te"],na.rm=T)
    
    #calculate daily ANP: compare current te value to grazed value from last sample period
    biomass.mean.x = dat2$totshtbiom.meanf[dat2$yr==yr&dat2$site==si&dat2$sample.==s&dat2$trt=="te"]
    days.x = dat2$days.1[dat2$yr==yr&dat2$site==si&dat2$sample.==s&dat2$trt=="te"]
    if(length(days.x)==0) days.x = 28 #if missing assume 28 d
    #if there is no te treatment, use gr instead (presumed no grazers around) and set dC to zero
    if(length(biomass.mean.x)==0) {biomass.mean.x = dat2$totshtbiom.meanf[dat2$yr==yr&dat2$site==si&dat2$sample.==s&dat2$trt=="gr"]; newdat$dC[i] = 0}
    if(length(biomass.mean.x)==0) next
    s1 = s-1 #last sample period
    grbiomass.last.mean.x = dat2$totshtbiom.meanf[dat2$yr==yr&dat2$site==si&dat2$sample.==s1&dat2$trt=="gr"]
    if(length(grbiomass.last.mean.x)==0) next
    newdat$dANP[i] = (biomass.mean.x - grbiomass.last.mean.x ) /  days.x
    
    #calculate consumption: compare current te value to current gr value
    biomass.mean.x = dat2$totshtbiom.meanf[dat2$yr==yr&dat2$site==si&dat2$sample.==s&dat2$trt=="te"]
    if(length(biomass.mean.x)==0) next
    grbiomass.mean.x = dat2$totshtbiom.meanf[dat2$yr==yr&dat2$site==si&dat2$sample.==s&dat2$trt=="gr"]
    if(length(grbiomass.mean.x)==0) next
    if(!is.na(newdat$dC[i])) next #if already filled above with zero, skip
    newdat$dC[i] = (biomass.mean.x - grbiomass.mean.x) / days.x
    
  } 
} #close sample loop

#convert daily to monthly values
newdat$mANP = newdat$dANP*28
newdat$mC = newdat$dC*28

```

Do these values make sense? Plotting C vs. ANP, with earlier season values darker gray (note, values at the bottom are those presumed to have zero consumption due to missing te treatment):

```{r,warning=F}

cols = gray.colors(7)[as.numeric(newdat$sample)]
par(mfrow=c(1,1),mar=c(5,5,1,1))
boxplot(mANP ~ sample,newdat)
boxplot(mC[mC!=0] ~ sample[mC!=0],newdat) #omit zeros to see boxes
plot(newdat$mANP+.1,newdat$mC+.1,log="xy",col=cols,cex=1.5,lwd=2,ylab="Monthly Consumption, g/m2",xlab="Monthly ANP, g/m2")
abline(0,1)
legend("topleft",legend=c(1:7),text.col=gray.colors(7))
```

This indicates that consumption values are higher for higher ANP values early in the growing season. Later in the year, consumption is often greater than ANP, suggesting old biomass is being consumed. Is this accurate?

And here is consumption as a (log) proportion of ANP, same colors as above. Consumption is proportionally low early in the growing season but much higher (approaching 100%) later in the season. As expected?

```{r}
#Consumption as a (log) proportion of ANP:
plot(newdat$mANP,newdat$mC/newdat$mANP,log="y",col=cols,cex=1.5,lwd=2,ylab="Log mC/mANP",xlab="Monthly ANP, g/m2")


```


### Modeling ANP and C

We can add predictors to the ANP-C dataframe to create explanatory models. 

```{r}
dat.env = datb[,c("yr","site","sample.","soilc","soiln","clay","silt","mo.ppt","mo.temp","elevation")]
  names(dat.env)[1:3] = c("year","site","sample")
nd = merge(newdat,dat.env,all.x=T,all.y=F)
nd = nd[!is.na(nd$mANP),] #exclude all rows with missing ANP data
nd = nd[duplicated(nd)==F,] #remove duplicate rows
```

#### Doug's Question 1. *How does aboveground production during the sampling periods (measured every 3-5 weeks) vary during the snow free season, from snowmelt to the last Sept-Oct sample date, and how do those dynamics vary with elevation, temp, precipitation, soil texture (i.e., clay content) and soil C?*

Here are a set of graphs that show monthly ANP for each sample period, for each site-year available. The first batch is is for 1988-1989 data:

<div style="background-color: #FFFF00">These graphs have been updated for absolute scaling of the Y axes (0-300 g/m2); the late season pulses don't seem as dramatic. </div>

```{r}

#We can have a look at every site for each year sampled:
tab = table(nd$site,nd$year)
tab

nd$sample = as.numeric(nd$sample)

#1988 and 1989
s = rownames(tab)[1:13]
par(mfrow=c(7,4),mar=rep(.6,4),oma=c(1,1,2,1))
for(i in 1:13) {
  site = s[i]
  if(length(nd$sample[nd$site==site&nd$year==1988])==0) plot(0,0,type="n",axes=F) else plot(nd$sample[nd$site==site&nd$year==1988],nd$mANP[nd$site==site&nd$year==1988],type="b",xlab="Sample",ylab="mANP",xlim=c(1,6),ylim=c(0,300),axes=F,xpd=T); mtext("",line=-1,cex=.5); mtext(site,side=2,cex=.8)
  if(i==1) mtext("1988",side=3,cex=1,line=0)
  if(i==2) mtext("1988",side=3,cex=1,line=0)
    if(length(nd$sample[nd$site==site&nd$year==1989])==0) plot(0,0,type="n",axes=F) else plot(nd$sample[nd$site==site&nd$year==1989],nd$mANP[nd$site==site&nd$year==1989],type="b",xlab="Sample",ylab="mANP",xlim=c(1,6),ylim=c(0,300),axes=F,xpd=T); mtext("",line=-1,cex=.5)
  if(i==1) mtext("1989",side=3,cex=1,line=0)
  if(i==2) mtext("1989",side=3,cex=1,line=0)
}

```

With only two exceptions, peak biomass is at the first sample period. Here is for sites 1999-2001, which suggest a more diverse set of responses:

```{r}

#1999, 2000, 2001
s = rownames(tab)[c(1,14:22)]
par(mfrow=c(5,3),mar=rep(.6,4),oma=c(1,1,2,1))
for(i in 1:5) {
  site = s[i]
  if(length(nd$sample[nd$site==site&nd$year==1999])==0) plot(0,0,type="n",axes=F) else plot(nd$sample[nd$site==site&nd$year==1999],nd$mANP[nd$site==site&nd$year==1999],type="b",xlab="Sample",ylab="mANP",xlim=c(1,7),ylim=c(0,300),axes=F,xpd=T); mtext("",line=-1,cex=.5); mtext(site,side=2,cex=.8)
  if(i==1) mtext("1999",side=3,cex=1,line=0)
    if(length(nd$sample[nd$site==site&nd$year==2000])==0) plot(0,0,type="n",axes=F) else plot(nd$sample[nd$site==site&nd$year==2000],nd$mANP[nd$site==site&nd$year==2000],type="b",xlab="Sample",ylab="mANP",xlim=c(1,7),ylim=c(0,300),axes=F,xpd=T); mtext("",line=-1,cex=.5)
  if(i==1) mtext("2000",side=3,cex=1,line=0)
      if(length(nd$sample[nd$site==site&nd$year==2001])==0) plot(0,0,type="n",axes=F) else plot(nd$sample[nd$site==site&nd$year==2001],nd$mANP[nd$site==site&nd$year==2001],type="b",xlab="Sample",ylab="mANP",xlim=c(1,7),ylim=c(0,300),axes=F,xpd=T); mtext("",line=-1,cex=.5)
  if(i==1) mtext("2001",side=3,cex=1,line=0)
}

par(mfrow=c(5,3),mar=rep(.6,4),oma=c(1,1,2,1))
for(i in 6:10) {
  site = s[i]
  if(length(nd$sample[nd$site==site&nd$year==1999])==0) plot(0,0,type="n",axes=F) else plot(nd$sample[nd$site==site&nd$year==1999],nd$mANP[nd$site==site&nd$year==1999],type="b",xlab="Sample",ylab="mANP",xlim=c(1,7),ylim=c(0,300),axes=F,xpd=T); mtext("",line=-1,cex=.5); mtext(site,side=2,cex=.8)
  if(i==6) mtext("1999",side=3,cex=1,line=0)
    if(length(nd$sample[nd$site==site&nd$year==2000])==0) plot(0,0,type="n",axes=F) else plot(nd$sample[nd$site==site&nd$year==2000],nd$mANP[nd$site==site&nd$year==2000],type="b",xlab="Sample",ylab="mANP",xlim=c(1,7),ylim=c(0,300),axes=F,xpd=T); mtext("",line=-1,cex=.5)
  if(i==6) mtext("2000",side=3,cex=1,line=0)
      if(length(nd$sample[nd$site==site&nd$year==2001])==0) plot(0,0,type="n",axes=F) else plot(nd$sample[nd$site==site&nd$year==2001],nd$mANP[nd$site==site&nd$year==2001],type="b",xlab="Sample",ylab="mANP",xlim=c(1,7),ylim=c(0,300),axes=F,xpd=T); mtext("",line=-1,cex=.5)
  if(i==6) mtext("2001",side=3,cex=1,line=0)
}

```

And finally here are the six sites for 2012-2014, which conform more to the 1988-89 patterns:

```{r}

#2012, 2013, 2014
s = rownames(tab)[c(1,23:27)]
par(mfrow=c(6,3),mar=rep(.6,4),oma=c(1,1,2,1))
for(i in 1:6) {
  site = s[i]
  if(length(nd$sample[nd$site==site&nd$year==2012])==0) plot(0,0,type="n",axes=F) else plot(nd$sample[nd$site==site&nd$year==2012],nd$mANP[nd$site==site&nd$year==2012],type="b",xlab="Sample",ylab="mANP",xlim=c(1,7),ylim=c(0,300),axes=F,xpd=T); mtext("",line=-1,cex=.5); mtext(site,side=2,cex=.8)
  if(i==1) mtext("2012",side=3,cex=1,line=0)
    if(length(nd$sample[nd$site==site&nd$year==2013])==0) plot(0,0,type="n",axes=F) else plot(nd$sample[nd$site==site&nd$year==2013],nd$mANP[nd$site==site&nd$year==2013],type="b",xlab="Sample",ylab="mANP",xlim=c(1,7),ylim=c(0,300),axes=F,xpd=T); mtext("",line=-1,cex=.5)
  if(i==1) mtext("2013",side=3,cex=1,line=0)
      if(length(nd$sample[nd$site==site&nd$year==2014])==0) plot(0,0,type="n",axes=F) else plot(nd$sample[nd$site==site&nd$year==2014],nd$mANP[nd$site==site&nd$year==2014],type="b",xlab="Sample",ylab="mANP",xlim=c(1,7),ylim=c(0,300),axes=F,xpd=T); mtext("",line=-1,cex=.5)
  if(i==1) mtext("2014",side=3,cex=1,line=0)
}

```

_Modeling at the monthly scale_. Because monthly climate have been provided, a first approach to addressing seasonal changes in ANP is using a monthly time step model, accounting for dependencies across sites within years, and across years within sites (crossed random effects of year and site). And we need an initial check of whether predictors are correlated. 

<div style="background-color: #FFFF00">Revised below models, excluding soil variables to increase sample size and adding rainfall-temp-elevation interactions. </div>


```{r}

#correlation matrix of predictors
cor(nd[,9:15],use = "pairwise.complete.obs")

#the only problem I see here is soilc and soiln:
plot(nd$soiln,nd$soilc)
#many soilc values are missing anyway, so we'll stick with soiln

#main effects only:
library(lme4)
library(lmerTest)
library(performance) #for nakagawa's R2
#lme1 = lmer(mANP ~ soiln + clay + silt + mo.ppt + mo.temp + elevation + (1|year) + (1|site), data=nd)
  #remove soil variables for now
lme1 = lmer(mANP ~ scale(mo.ppt)*scale(mo.temp)*scale(elevation) + (1|year) + (1|site), data=nd)
summary(lme1)
r2_nakagawa(lme1) #32% R2

  #lots going on here

```

<div style="background-color: #FFFF00"> Comparing above model without sample 1 values. </div>


```{r}

#model above, ignoring sample 1 values
lme2 = lmer(mANP ~ scale(mo.ppt)*scale(mo.temp)*scale(elevation) + (1|year) + (1|site), data=nd, subset=nd$sample!=1)
summary(lme2)
r2_nakagawa(lme2)
  #interestingly, much reduced R2, otherwise similar

#graphing to show sample 1 values
par(mfrow=c(1,1),mar=c(5,5,1,1))
cols = gray.colors(7)[as.numeric(nd$sample)]
plot(nd$mo.ppt,nd$mANP,xlab="Monthly rainfall, mm",ylab="Monthly ANP, g/m2",col=cols)
points(nd$mo.ppt[nd$sample==1],nd$mANP[nd$sample==1],col="blue")
plot(nd$silt,nd$mANP,xlab="Silt %",ylab="Monthly ANP, g/m2",col=cols)
points(nd$silt[nd$sample==1],nd$mANP[nd$sample==1],col="blue")
```

<div style="background-color: #FFFF00"> Considering %dead biomass alongside rainfall effect. Darker symbols indicate more % dead biomass. </div>

```{r}
cols2 = gray.colors(10)[rev(as.numeric(cut(nd$percDead,10)))]
plot(nd$mo.ppt[nd$sample!=1],nd$mANP[nd$sample!=1],xlab="Monthly rainfall, mm",ylab="Monthly ANP, g/m2",col=cols2[nd$sample!=1],cex=1.5)

```


#### Doug's Question 2. *How does grazing intensity affect aboveground production during sampling periods and how do those animal effects interact with the climate and site factors?*

```{r}
nd$pmC = nd$mC / nd$mANP #consumption as a proportion of ANP
summary(nd$pmC)
nd$pmC[nd$pmC==-Inf] = NA #convert div0 to NA
nd$pmC[nd$pmC==Inf] = NA #convert div0 to NA

#should values below 0 or above 1 be considered or removed? these are cases where grazers ate more than estimated ANP
#for now, I'll change to NA
nd$pmC[nd$pmC<0 | nd$pmC > 1] = NA
summary(nd$pmC)
length(nd$pmC) #44% are NA! huge loss of data

#let's keep a version of pmC that retains values below 0 and above 1
nd$pmC2 = nd$mC / nd$mANP #consumption as a proportion of ANP
nd$pmC2[nd$pmC2==-Inf] = NA #convert div0 to NA
nd$pmC2[nd$pmC2==Inf] = NA #convert div0 to NA
summary(nd$pmC2) #only 11 NAs

#create a column that, for each row, is the pmC of the previous sample period
nd$pmC1 = NA #fill column
nd$pmC1v2 = NA #fill column, version that includes below 0 and above 1 values
for(i in 1:dim(nd)[1]) {
  if(nd$sample[i]==1) next #ignore sample period 1
  si = nd$site[i]; yr = nd$year[i]; sa = nd$sample[i] - 1
  if(length(nd$pmC[nd$site==si&nd$year==yr&nd$sample==sa])==0) next
  nd$pmC1[i] = nd$pmC[nd$site==si&nd$year==yr&nd$sample==sa]
  if(length(nd$pmC[nd$site==si&nd$year==yr&nd$sample==sa])==0) next
  nd$pmC1v2[i] = nd$pmC2[nd$site==si&nd$year==yr&nd$sample==sa]
}
summary(nd$pmC1)
summary(nd$pmC1v2)

#note some duplicate values for site env factors in nd: can Doug check these?
```

<div style="background-color: #FFFF00"> Model changes: removing soil variables, adding %dead biomass. </div>


```{r}

#Building off of lme3, and including previous sample period grazing intensity:
lme4 = lmer(mANP ~ scale(mo.ppt) * scale(pmC1) * scale(percDead) + (1|year) + (1|site), data=nd,subset=nd$sample!=1)
summary(lme4)

#include pmC1v2 as driver with more data:
lme5 = lmer(mANP ~ scale(mo.ppt) * scale(pmC1v2) * scale(percDead) + (1|year) + (1|site), data=nd,subset=nd$sample!=1)
summary(lme5)

plot(nd$pmC1[nd$sample!=1],nd$mANP[nd$sample!=1],xlab="Proportion consumed of prior month",ylab="Monthly ANP, g/m2",col=cols2[nd$sample!=1],cex=1.5)

```

<div style="background-color: #FFFF00">

Still no evidence for previous-month consumption effects, although percentage dead biomass is a strong predictor and interacts with rainfall. I think it would be wise for Doug to compare these data to previous analyses showing a grazing priming effect--there may be something subtly different about this analysis.

</div>

#### Doug's Question 3. *How is annual aboveground production (ANPP) determined by climate (different growing season ppt and water year ppt amounts measured during different years), elevation, and site soil properties (texture, C)?*

Here I've created an annual dataset by using the provided annual climate variables and summing ANP across sample periods. This gives a sample size of about 60 site-years. 

<div style="background-color: #FFFF00">I've also created a new ANPP dependent variable that excludes monthly negative ANP values, called ANPP2. </div>

```{r}
#summarize to annual
library(doBy)
sumf = function(x) {sum(x,na.rm=T)}
ndA = summaryBy (mANP + mC ~ year + site, data = newdat, FUN = sumf )
  names(ndA)[c(3,4)] = c("ANPP","annualC")
ndA = ndA[ndA$ANPP>0,] #only 60-odd site-years available, omit NAs
hist(ndA$ANPP,breaks=20) 

#sum mANP values but ignore negative values
sumf2 = function(x) {sum(x[x>=0],na.rm=T)}
ndA2 = summaryBy (mANP ~ year + site, data = newdat, FUN = sumf2 )
names(ndA2)[3] = "ANPP2"
ndA = merge(ndA,ndA2,all.x=T)

#examine relationship of ANPP and ANPP2
plot(ndA$ANPP,ndA$ANPP2); abline(0,1)

#create another ANPP variable that only totals ANP for periods 2-7 (ignoring negatives)
newdat.no1 = newdat[newdat$sample!=1,] #take out all sample = 1 values
ndA2 = summaryBy (mANP ~ year + site, data = newdat.no1, FUN = sumf2 )
names(ndA2)[3] = "ANPP.no1"
ndA = merge(ndA,ndA2,all.x=T)

#examine relationship of ANPP2 and ANPP.no1
plot(ndA$ANPP2,ndA$ANPP.no1); abline(0,1)

#create new variable that is sample 1 mANP as a proportion of total (no negatives)
ndA$s1prop = (ndA$ANPP2-ndA$ANPP.no1) / ndA$ANPP2
hist(ndA$s1prop)

#merge annual and site values
dat.env2 = datb[,c("yr","site","soilc","soiln","clay","silt","wateryr.ppt","gsyr.ppt","elevation","mo.temp")]
  names(dat.env2)[1:2] = c("year","site")
ndA2 = merge(ndA,dat.env2,all.x=T,all.y=F)  
ndA2 = ndA2[duplicated(ndA2)==F,] 
dim(ndA2)  

#some site-years have multiple values for soil texture; simplify to mean
meanf = function(x) {mean(x,na.rm=T)}
ndA3 = summaryBy (. ~ year + site, data = ndA2, FUN = meanf , keep.names=T)
str(ndA3)
names(ndA3)[15] = "mean.temp" #added mean gs temperature by taking monthly means
summary(ndA3$clay) #about half missing
summary(ndA3$silt) #about half missing
summary(ndA3$soilc) #25 missing
summary(ndA3$soiln)
dim(ndA3)
```

Here is an annual modeling approach similar to that above:

<div style="background-color: #FFFF00">Models have been revised: </div>

```{r}
cor(ndA3[,8:15],use="pairwise.complete.obs")
  #omit soilc again, and see whether gsyr.ppt or wateryr.ppt is a better precip variable
cor(ndA3[,c(5,12,13)])
  #slightly higher correlation of ANPP with gsyr.ppt

library(lme4)
#lme1 = lmer(ANPP ~ scale(silt)*scale(gsyr.ppt)*scale(elevation) + (1|year) + (1|site), data=ndA3)
  #soil variables have too many NAs, so ignore for now

lme1 = lmer(ANPP2 ~ scale(gsyr.ppt)*scale(elevation)*scale(mean.temp) + (1|year) + (1|site), data=ndA3)
summary(lme1)
r2_nakagawa(lme1)
  #no main effects, but sig positive interaction of elevation and mean temp

#compare to model without random effects
lm1 = lm(ANPP2 ~ scale(gsyr.ppt)*scale(elevation)*scale(mean.temp), data=ndA3)
summary(lm1)
  #no relationships
  #doesn't matter if ANPP or ANPP2 y variables are used

#compare to model without sample 1 values
lme2 = lmer(ANPP.no1 ~ scale(gsyr.ppt)*scale(elevation)*scale(mean.temp) + (1|year) + (1|site), data=ndA3)
summary(lme2)
  #in this case, elevation has significant negative correlation with ANPP
  #small neg effect of mean temp
```

<div style="background-color: #FFFF00">Model results have changed, after excluding soil factors and including temperature: </div>

Excluding the soil variables allows for a much larger dataset (N>60) but neither growing season precip nor elevation, nor their interaction, explains total ANPP (measured with or without monthly negative values), *although there is a small positive interaction of elevation and temp, which makes sense*. However, if the first month of ANP values are excluded, then ANPP decreases with elevation and (marginally) temperature.
  
#### Doug's Question 4. *How does grazing determine ANPP and interact with the same climate (growing season, water yr) and site (elevation, soil) factors. 

I'll assume that we use the same model as for 3, but also include the ANPP estimates for the pe treatment (estimated as peak biomass). 

<div style="background-color: #FFFF00"> Below I used ANPP2 rather than ANPP, which ignores negative monthly ANP values. I also added ANPP to the model explaining the difference between grazed and ungrazed grasslands. It has a large positive effect, as suggested by the below graph.

</div>

```{r}
#grab pe treatments
dpe = dat[dat$trt=="pe",]

#simplify to mean biomass per sample period
meanf = function(x) {mean(x,na.rm=T)}
sef = function(x) {sd(x,na.rm=T) / sqrt(length(x))}
dpe2 = summaryBy( totshtbiom ~ yr + site + sample. + days.1 + juldate, data = dpe, FUN=c(meanf,sef) , keep.names=T)
str(dpe2)
table(dpe2$site,dpe2$yr)

#assume ANPP is estimated by peak (max) biomass
maxf = function(x) {max(x,na.rm=T)}
dpe3 = summaryBy( totshtbiom.meanf ~ yr + site, data = dpe2, FUN=c(maxf), keep.names=T)
names(dpe3)[c(1,3)] = c("year","ANPP.pe")
par(mfrow=c(1,1))
boxplot(ANPP.pe ~ site + year, dpe3) #range of values for 39 site-years for pe treatment

#add pe values to ndA3 dataset
annual = merge(ndA3,dpe3,all.x=T)

```

Here is the relationship between ANPP on grazed lands vs. ANPP on permanent exclosures, darker colors are years of more rainfall.

```{r}
cols = gray.colors(20)[cut(annual$gsyr.ppt,8)]
plot(annual$ANPP.pe,annual$ANPP,col=cols,pch=19,cex=1.5)
abline(0,1)

plot(annual$ANPP.pe,annual$ANPP2,col=cols,pch=19,cex=1.5)
abline(0,1)
abline(lsfit(annual$ANPP.pe,annual$ANPP2),lwd=2)
title(main="ANPP excl. negative mon values")
  #possibly a larger difference in higher ANPP sites

```

I'll make a new dependent variable that is the difference between ANPP and ANPP.pe, and model this as we did above.

```{r}

annual$Adiff = annual$ANPP - annual$ANPP.pe #positive values indicate grazer stimulation
hist(annual$Adiff)
#note many missing silt values; ignoring for now

#version ignoring negative monthly ANP values
annual$Adiff2 = annual$ANPP2 - annual$ANPP.pe #positive values indicate grazer stimulation


library(lme4)
#lme1 = lmer(Adiff ~ scale(gsyr.ppt) + scale(soiln) + scale(elevation) + (1|year) + (1|site), data=annual)
#summary(lme1)
  #no effects

#new model:
lme1 = lmer(Adiff2 ~ scale(ANPP2)*scale(gsyr.ppt) * scale(elevation) * scale(mean.temp) + (1|year) + (1|site), data=annual)
summary(lme1)
  #positive ANPP effect, moderated by rainfall  

```



